### 2023-06-25 Draft.tex

Note that for Corner filtering, the kernel is not normalized. Normalization is important as otherwise, a pixel may go beyond the allowed value range.
In cases where the kernel has a negative value, it means that the pixel value is inverted, since an integer overflow occurs (e.g. $64
(-1)= -64 \mod 256 = 192$ for one channel in an 8 bit pixel).
With some filters, it is not always possible to separate it into just two operations, therefore it can be
transformed into a series of 1 dimensional operations. However, this may be impractical as it increases the time and space
complexity.

\subsection{Examples of Linear Filters}
\begin{itemize}
    \item \textbf{Box Filter:} each pixel is determined by the average of a $K \times K$ 
        neighborhood of pixels. It is also often called the \emph{moving average}, as the calculation can be done with scanlines, subtracting the
        old pixel and adding the newest pixels to the running sum. As the pixel is influenced equally by a square neighborhood, this may not be ideal as some `square artifacts' can be seen.
    \item \textbf{Bartlett Filter:} one solution to this is by separably convolving the image
        with a linear `tent' function (having its peak in the middle). A $3\times 3$ version of this is called the bilinear kernel.
    \item \textbf{Approximate Gaussian Filter:} if the Bartlett kernel is convolved with itself, this will produce an approximate version of the gaussian kernel.
\end{itemize}

@@

\subsection{Band-pass and steerable Filter}
In the above example, the sobel filter is calculated either for the horizontal or
vertical axis. Nevertheless, it is possible to calculate filters (including for sobel) for any angles in between using a specified 
\emph{directional derivative} $\nabla_{\boldsymbol{\hat{u}}} = \frac{\partial}{\partial\boldsymbol{\hat{u}}}$ 
and a \emph{unit direction} $\boldsymbol{\hat{u}} = (\cos \theta, \sin \theta)$\cite{Szeliski_2022}.
\[
    \boldsymbol{\hat{u}} \cdot \nabla (G \ast f) = \nabla_{\boldsymbol{\hat{u}}}(G \ast f) = (\nabla_{\boldsymbol{\hat{u}}}G) \ast  f
\]
Then it is also preferable to smooth it with the Gaussian filter.
\[
    G_{\boldsymbol{\hat{u}}} = u G_x + v G_y = u \frac{\partial G}{\partial x} + v \frac{\partial G}{\partial y} 
\]
This is called the \emph{directional} or \emph{oriented} filter.
Another filter is called the \emph{Laplacian operator}, which is the second (undirected)derivative.\cite{Szeliski_2022}
\[
    \nabla^2 f = \frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}
\]
and again, it can be smoothed with the Gaussian filter. It is equivalent to the \emph{Laplacian of Gaussian}(LoG) filter.
\[
    \nabla^2 G (x, y, \sigma) = (\frac{x^2 + y^2}{\sigma^4} - \frac{2}{\sigma^2}) G (x, y, \sigma)
\]
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacianOriginal.png}}
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacian.png}}
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacianAdded.png}}
    \caption{Laplacian Filtering\cite{NASA_on_The_Commons_1970} (a) Orinal image of NASA pilot; (b) Output of the laplacian filter with Gaussian blur (c) Combined image of the orientation map and original}
    \label{fig:padding_borders}
\end{figure}\end{minipage}

>>>>

Note that for Corner filtering, the kernel is not normalized. Normalization is important as otherwise, a pixel may go beyond the allowed value range.
In cases where the kernel has a negative value, it means that the pixel value is inverted, since an integer overflow occurs (e.g. $64
(-1)= -64 \mod 256 = 192$ for one channel in an 8 bit pixel).`'
With some filters, it is not always possible to separate it into just two operations, therefore it can be
transformed into a series of 1 dimensional operations. However, this may be impractical as it increases the time and space
complexity.

\subsection{Examples of Linear Filters}
Due to the fact that linear filter in essence is the sum of pixels in a neighborhood with varying weights and sizes, there are infinitely many different kernels that can
define operators which classified as linear and invariant. It is also even possible to factor in different color channels. As an example, it is possible to shift the hue of an image.
In this example, using a kernel of $\begin{bmatrix}0 & 0 & 1\end{bmatrix}$, the hue is shifted to the left (negative).
\lstset{language=Python}
\begin{lstlisting}
img = cv.imread("color.jpg")
kernel = np.array([[[0, 0, 1]]])
convolved = ndimage.convolve(img, kernel, mode='reflect')
cv.imwrite('../img/colorShift.jpg', convolved)
\end{lstlisting}

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.30\textwidth]{codes/color.jpg}}
    \subfigure[shifted]{\includegraphics[width=0.30\textwidth]{img/colorShift.jpg}}
    \caption{Hue shifting using linear filters of the GALEX telescope image\cite{NASA_on_The_Commons_23}}
    \label{fig:hueshift}
\end{figure}\end{minipage}

The simplest application for linear filters are for smoothing(blurring) images. This not only serves their inherent purpose, but as later discussed, it can be used to aid other applications such as
edge detection or image enhancement.
\begin{itemize}
    \item \textbf{Box Filter:} each pixel is determined by the average of a $K \times K$ 
        neighborhood of pixels. It is also often called the \emph{moving average}, as the calculation can be done with scanlines, subtracting the
        old pixel and adding the newest pixels to the running sum. As the pixel is influenced equally by a square neighborhood, this may not be ideal as some `square artifacts' can be seen.
    \item \textbf{Bartlett Filter:} one solution to this is by separably convolving the image
        with a linear `tent' function (having its peak in the middle). A $3\times 3$ version of this is called the bilinear kernel.
    \item \textbf{Approximate Gaussian Filter:} if the Bartlett kernel is convolved with itself, this will produce an approximate version of the gaussian kernel.
\end{itemize}

This intuitively can be used to reduce the amount of noise on a picture. However, since it naively blurs the entire image,
details are also lost in the process.\\
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.32\textwidth]{codes/original_noise.jpg}}
    \subfigure[noise]{\includegraphics[width=0.32\textwidth]{codes/noise.jpg}}
    \subfigure[gaussian]{\includegraphics[width=0.32\textwidth]{img/noiseSuppressed.jpg}}
    \caption{(a) Original image of the Star City Museum\cite{NASA_on_The_Commons_14}(b) Noise added artificially (c) Blurred with Gaussian (k=27 $\sigma=7$)}
    \label{fig:noise}
\end{figure}\end{minipage}

@@

\subsection{Band-pass and steerable Filter}
In the above example, the sobel filter is calculated either for the horizontal or
vertical axis. Nevertheless, it is possible to calculate filters (including for sobel) for any angles in between using a specified 
\emph{directional derivative} $\nabla_{\boldsymbol{\hat{u}}} = \frac{\partial}{\partial\boldsymbol{\hat{u}}}$ 
and a \emph{unit direction} $\boldsymbol{\hat{u}} = (\cos \theta, \sin \theta)$\cite{Szeliski_2022}.
\[
    \boldsymbol{\hat{u}} \cdot \nabla (G \ast f) = \nabla_{\boldsymbol{\hat{u}}}(G \ast f) = (\nabla_{\boldsymbol{\hat{u}}}G) \ast  f
\]
Then it is also preferable to smooth it with the Gaussian filter.
\[
    G_{\boldsymbol{\hat{u}}} = u G_x + v G_y = u \frac{\partial G}{\partial x} + v \frac{\partial G}{\partial y} 
\]
This is called the \emph{directional} or \emph{oriented} filter.

A similar to operator to the sobel filter is called the \emph{Prewitt operator}, which instead of convolving it with a perpendicular tent filter,
it instead uses an average operator. Commonly however, the Sobel operator is preferred as it provides better performance\cite{NIXON2020141}. Through testing, the Sobel Operator seems
to be less selective than the Prewitt Operator, but one can argue that the Prewitt is then preferred as it is more accurate.

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[Pattern]{\includegraphics*[width=0.29\textwidth]{codes/pattern.png}}
    \subfigure[Sobel]{\includegraphics[width=0.29\textwidth]{img/directionSobel.png}}
    \subfigure[Prewitt]{\includegraphics[width=0.29\textwidth]{img/directionScharr.png}}
    \caption{Vertical edge detection using Sobel and Prewitt on a test pattern}\label{fig:pattern}
\end{figure}\end{minipage}

As seen on Figure \ref{fig:pattern}, the Sobel filter tapers the edges at the top into a gradient, whereas the Prewitt marks it confidently without tapering. In practice however,
the minimal difference might not be worth the difference in the additional compute time \cite{NIXON2020141}.

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[]{
        {$\begin{bmatrix}
            -1 & 0 & 1 \\
            -2 & 0 & 2 \\
            -1 & 0 & 1
        \end{bmatrix}$}
    }
    \subfigure[]{
        {$\begin{bmatrix}
            1 & 0 & -1 \\
            1 & 0 & -1 \\
            1 & 0 & -1
        \end{bmatrix}$}
    }
    \subfigure[]{
        {$\begin{bmatrix}
            1 & 1 & 1 \\
            0 & 0 & 0 \\
            -1 & -1 & -1
        \end{bmatrix}$}
    }
    \caption{Sobel and Prewitt operator comparison (a) Horizontal Sobel operator; 
    (b) Horizontal Prewitt operator; (c) Vertical Prewitt operator}\label{fig:prewittSobel}
\end{figure}\end{minipage}

Another filter is called the \emph{Laplacian operator}, which is the second (undirected)derivative.\cite{Szeliski_2022}
\[
    \nabla^2 f = \frac{\partial^2 f}{\partial x^2} + \frac{\partial^2 f}{\partial y^2}
\]
and again, it can be smoothed with the Gaussian filter. It is equivalent to the \emph{Laplacian of Gaussian}(LoG) filter.
\[
    \nabla^2 G (x, y, \sigma) = (\frac{x^2 + y^2}{\sigma^4} - \frac{2}{\sigma^2}) G (x, y, \sigma)
\]
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacianOriginal.png}}
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacian.png}}
    \subfigure[]{\includegraphics[width=0.29\textwidth]{img/laplacianAdded.png}}
    \caption{Laplacian Filtering (a) Original image of NASA pilot\cite{NASA_on_The_Commons_1970}; (b) Output of the laplacian filter with Gaussian blur (c) Combined image of the orientation map and original}
    \label{fig:padding_borders}
\end{figure}\end{minipage}
